---
title: "Extracting stream and catchment geometry data for the Goulburn River"
author: "Thomas Wilkins"
format: html
editor: visual
---

```{r}
#| label: setup
#| echo: false
#| message: false
#| warning: false

source("functions/misc_functions.R")

load_packages(c(
  "sf",
  "dplyr",
  "ggplot2",
  "igraph"
))

# set path to geofabric data
geofab_network_path <- "data/SH_Network_GDB/"
geofab_catchments_path <- "data/HR_Catchments_GDB/"


```

## 1. Introduction

This document contains R code to extract stream and catchment geometry data for the Goulburn River and related territories.
Stream and catchment data was extracted from the Australian Hydrological Geospatial Fabric (Geofabric).

There are two layers of interest in this dataset:

1.  AHGFNetworkStream: contains stream network geometry and attributes
2.  AHGFCatchment: contains catchment geometry and attributes

### 1.1. Extracting streams connected to the Murray River outlet

Initial exploration was done to identify all streams connected to the Murray River outlet, as the Goulburn River is a tributary of the Murray River.
The most downstream point of the Murray River was identified as the Coorong Channel outlet to the Southern Ocean, near Goolwa, South Australia.

```{r}
#| label: extract_geofab_streams_data
#| echo: true
#| message: false
#| warning: false
#| results: 'hide'


st_layers(paste0(geofab_network_path, "SH_Network.gdb/"))

# read in geofabric stream network data
geofab_networkstream <- sf::st_read(
  dsn = paste0(geofab_network_path, "SH_Network.gdb/"),
  layer = "AHGFNetworkStream"
)

# Isolate most downstream point of the Murray River:
murray_ol <- geofab_networkstream[geofab_networkstream$Name == "COORONG CHANNEL" & 
                                    geofab_networkstream$NextDownID == -1, ] # -1 implies outlet
murray_ol <- murray_ol[!is.na(murray_ol$NextDownID), ] 

# Create edge list (upstream → downstream)
edges <- geofab_networkstream %>%
  st_drop_geometry() %>%
  select(From_Node, To_Node, HydroID)

# Create graph object of the stream network
g <- graph_from_data_frame(edges, directed = TRUE)

# Find all upstream HydroIDs of the Murray River outlet
upstream_nodes <- subcomponent(g, as.character(murray_ol$To_Node), mode = "in")
streams_murray_connected <- geofab_networkstream %>%
  filter(From_Node %in% names(V(g))[upstream_nodes] |
         To_Node %in% names(V(g))[upstream_nodes])

# Compute shortest paths *upstream* (reverse direction)
# Add distances to streams_murray_connected
streams_murray_connected <- streams_murray_connected %>%
  mutate(distance_to_murray_ol = st_distance(Shape, st_geometry(murray_ol)))
  mutate(distance_to_murray_ol = as.numeric(distance_to_murray_ol)) # convert to numeric

streams_murray_connected$d2ol <- as.numeric(streams_murray_connected$distance_to_murray_ol)

# Identify confluences
# A confluence is where two or more upstream segments flow into a single downstream segment
# i.e., a To_Node with multiple From_Nodes
streams_murray_connected$confluence <- 0
to_node_counts <- as.data.frame(table(streams_murray_connected$To_Node))
confluence_nodes <- to_node_counts$Var1[to_node_counts$Freq > 1]
streams_murray_connected$confluence[streams_murray_connected$To_Node %in% confluence_nodes] <- 1
streams_murray_confluences <- streams_murray_connected[streams_murray_connected$confluence == 1 & 
                                                         streams_murray_connected$Name == " ", ]

# Save streams_murray_connected for later use
saveRDS(streams_murray_connected, "data/murray_networkstream.rds")


```

The following plot shows all streams connected to the Murray River outlet, overlaid on a map of Australia for verification.

```{r}

# Read murray stream network data:
streams_murray_connected <- readRDS("data/murray_networkstream.rds")

# Extract coordinates of the most downstream point of the Murray River (murray_ol)
murray_ol <- streams_murray_connected[streams_murray_connected$Name == "COORONG CHANNEL" & 
                                    streams_murray_connected$NextDownID == -1, ]

murray_ol_geom <- st_geometry(murray_ol)[[1]]
coords <- st_coordinates(murray_ol_geom)
ds_point <- coords[nrow(coords), ]

# Convert to sf POINT
murray_ol_sf <- st_as_sf(data.frame(x = ds_point[1], y = ds_point[2]),
                       coords = c("x", "y"),
                       crs = st_crs(geofab_networkstream))


# Plot streams_murray_connected overlaid on a map of Australia to verify
australia <- rnaturalearth::ne_countries(scale = "medium", returnclass = "sf")
ggplot() +
  geom_sf(data = australia, fill = "lightgrey") +
  geom_sf(data = streams_murray_connected, color = "blue") +
  # Plot a point for the downstream-most point on the Murray River (murray_ol):
  geom_sf(data = murray_ol_sf, color = "red", size = 2) +
  coord_sf(xlim = c(138, 155), ylim = c(-40, -25)) +
  theme_minimal() +
  labs(title = "Streams connected to the Murray River outlet")


```

### 1.2. Extracting catchments within the Murray-Darling Basin

Catchments were derived from the Geofabric HR_Catchments geodatabase.

```{r}
#| label: extract_geofab_streams_data
#| echo: true
#| message: false
#| warning: false
#| results: 'hide'

st_layers(paste0(geofab_network_path, "SH_Network.gdb/"))



```

## Missing Stream names

Approximately 60% of the stream segments in the Geofabric database have no stream name assigned.
Correcting the whole Murray-Darling basin is beyond the scope of this document (for now), so I will focus on correcting the Goulburn River stream names.
The Goulburn River has several segments in the Geofabric data, some of which are unnamed.
We can identify these segments based on their proximity to known locations along the Goulburn River.
Given that each stream has a downstream confluence to a distinct stream, and an upstream-most point, we can use these characteristics to identify the Goulburn River segments if a stream has at least one named component.
Once this is achieved in the Goulburn River, it can be extended to the full Murray-Darling stream network.

First, I need to find the downstream-most point of the Goulburn River catchment, which (I presume) is the confluence with the Murray River near Echuca, Victoria.

```{r}
#| label: goulburn_name_correction
#| echo: true
#| message: false
#| warning: false
#| results: 'hide'

# Identify Goulburn River segments based on known named segments
goulburn_streams <- streams_murray_connected[streams_murray_connected$Name %in% c("GOULBURN RIVER", "GOULBURN WEIR TO SHEPPARTON"),]

# Just need one segment - most downstream segment is sufficient
goulburn_segment <- goulburn_streams[which.min(goulburn_streams$d2ol), ]

# find all upstream segments connected to the identified Goulburn River segment
goulburn_ig <- subcomponent(g, as.character(goulburn_segment$From_Node), mode = "in")

goulburn_us <- streams_murray_connected %>%
  filter(From_Node %in% names(V(g))[goulburn_ig] |
         To_Node %in% names(V(g))[goulburn_ig])


# Plot streams_murray_connected overlaid on a map of Australia to verify
australia <- rnaturalearth::ne_countries(scale = "medium", returnclass = "sf")
ggplot() +
  geom_sf(data = australia, fill = "lightgrey") +
  geom_sf(data = goulburn_us, color = "blue") +
  # Plot a point for the downstream-most point on the Murray River (murray_ol):
  # geom_sf(data = murray_ol_sf, color = "red", size = 2) +
  coord_sf(xlim = c(143, 149), ylim = c(-39, -34)) +
  theme_minimal() +
  labs(title = "Streams connected to the Murray River outlet")



# Find all downstream segments connected to the identified Goulburn River segment
downstream_nodes <- subcomponent(g, as.character(goulburn_streams$To_Node), mode = "out")

goulburn_ds <- streams_murray_connected %>%
  filter(From_Node %in% names(V(g))[downstream_nodes] |
         To_Node %in% names(V(g))[downstream_nodes])

```
