---
title: "Extracting stream and catchment geometry data for the Goulburn River"
author: "Thomas Wilkins"
format: html
editor: visual
bibliography: references.bib
---

```{r}
#| label: setup
#| echo: false
#| message: false
#| warning: false

source("functions/misc_functions.R")

load_packages(c(
  "sf",
  "dplyr",
  "ggplot2",
  "igraph"
))

# set path to geofabric data
geofab_network_path <- "data/SH_Network_GDB/"
geofab_catchments_path <- "data/HR_Catchments_GDB/"


```

## 1. Introduction

This document contains R code to extract stream and catchment geometry data for the Goulburn River and related territories.
The purpose of this document is to provide a clear and reproducible method for obtaining the necessary spatial data for hydrological analysis and modelling, specifically for the Flow-MER project.

Stream and catchment data was extracted from the Australian Hydrological Geospatial Fabric (Geofabric).

There are two layers of interest in this dataset:

1.  AHGFNetworkStream: contains stream network geometry and attributes
2.  AHGFCatchment: contains catchment geometry and attributes

## 2. Deriving a stream network for the Murray-Darling Basin, and Goulburn River and Northern Tributaries

### 2.1. Extracting streams connected to the Murray River outlet

Initial exploration was done to identify all streams connected to the Murray River outlet, as the Goulburn River is a tributary of the Murray River.
The most downstream point of the Murray River was identified as the Coorong Channel outlet to the Southern Ocean, near Goolwa, South Australia.

```{r}
#| label: extract_geofab_streams_data
#| echo: true
#| message: false
#| warning: false
#| results: 'hide'


st_layers(paste0(geofab_network_path, "SH_Network.gdb/"))

# read in geofabric stream network data
geofab_networkstream <- sf::st_read(
  dsn = paste0(geofab_network_path, "SH_Network.gdb/"),
  layer = "AHGFNetworkStream"
)

# Isolate most downstream point of the Murray River:
murray_ol <- geofab_networkstream[geofab_networkstream$Name == "COORONG CHANNEL" & 
                                    geofab_networkstream$NextDownID == -1, ] # -1 implies outlet
murray_ol <- murray_ol[!is.na(murray_ol$NextDownID), ] 

# Create edge list (upstream → downstream)
edges <- geofab_networkstream %>%
  st_drop_geometry() %>%
  select(From_Node, To_Node, HydroID)

# Create graph object of the stream network
g <- graph_from_data_frame(edges, directed = TRUE)

# Find all upstream HydroIDs of the Murray River outlet
upstream_nodes <- subcomponent(g, as.character(murray_ol$To_Node), mode = "in")
streams_murray_connected <- geofab_networkstream %>%
  filter(From_Node %in% names(V(g))[upstream_nodes] |
         To_Node %in% names(V(g))[upstream_nodes])

# Compute shortest paths *upstream* (reverse direction)
# Add distances to streams_murray_connected
streams_murray_connected <- streams_murray_connected %>%
  mutate(distance_to_murray_ol = st_distance(Shape, st_geometry(murray_ol)))
  mutate(distance_to_murray_ol = as.numeric(distance_to_murray_ol)) # convert to numeric

streams_murray_connected$d2ol <- as.numeric(streams_murray_connected$distance_to_murray_ol)

# Identify confluences
# A confluence is where two or more upstream segments flow into a single downstream segment
# i.e., a To_Node with multiple From_Nodes
streams_murray_connected$confluence <- 0
to_node_counts <- as.data.frame(table(streams_murray_connected$To_Node))
confluence_nodes <- to_node_counts$Var1[to_node_counts$Freq > 1]
streams_murray_connected$confluence[streams_murray_connected$To_Node %in% confluence_nodes] <- 1
streams_murray_confluences <- streams_murray_connected[streams_murray_connected$confluence == 1 & 
                                                         streams_murray_connected$Name == " ", ]

# Save streams_murray_connected for later use
saveRDS(streams_murray_connected, "data/murray_networkstream.rds")

### NAMED network
streams_murray_named <- streams_murray_connected[!streams_murray_connected$Name %in% c(""," "), ]
saveRDS(streams_murray_named, "data/murray_networkstream_named.rds")


```

The following plot shows all streams connected to the Murray River outlet, overlaid on a map of Australia for verification.

```{r}
#| label: fig-plot_murray_streams
#| echo: true
#| message: false
#| warning: false
#| results: 'hide'
#| fig-cap: "Streams connected to the Murray River outlet"
#| fig-width: 8
#| fig-height: 6

# Read murray stream network data:
streams_murray_connected <- readRDS("data/murray_networkstream.rds")
streams_murray_named <- readRDS("data/murray_networkstream_named.rds")
streams_murray_unnamed <- streams_murray_connected[streams_murray_connected$Name %in% c(""," "), ]

# Extract coordinates of the most downstream point of the Murray River (murray_ol)
murray_ol <- streams_murray_connected[streams_murray_connected$Name == "COORONG CHANNEL" & 
                                    streams_murray_connected$NextDownID == -1, ]

murray_ol_geom <- st_geometry(murray_ol)[[1]]
coords <- st_coordinates(murray_ol_geom)
ds_point <- coords[nrow(coords), ]

# Convert to sf POINT
murray_ol_sf <- st_as_sf(data.frame(x = ds_point[1], y = ds_point[2]),
                       coords = c("x", "y"),
                       crs = st_crs(geofab_networkstream))


# Plot streams_murray_connected overlaid on a map of Australia to verify
australia <- rnaturalearth::ne_countries(scale = "medium", returnclass = "sf")
ggplot() +
  geom_sf(data = australia, fill = "lightgrey") +
  
  # Create stream lines for streams_murray_connected, with a thin line:
#  geom_sf(data = streams_murray_unnamed, color = "lightblue1", linewidth = 0.5) +
  geom_sf(data = streams_murray_named, color = "blue", linewidth = 0.5) +
  # Plot a point for the downstream-most point on the Murray River (murray_ol):
  geom_sf(data = murray_ol_sf, color = "red", size = 2) +
  coord_sf(xlim = c(138, 155), ylim = c(-40, -25)) +
  theme_minimal() +
  labs(title = "Streams connected to the Murray River outlet")


```

### 2.2. Filtering streams within the Goulburn River and Northern Tributaries (GRANT) area

For the Flow-MER project, we are interested in streams within the Goulburn River and Northern Tributaries (GRANT) area.
This area includes the Goulburn River and its major tributaries: Loddon River, Broken Creek, Broken River, and Campaspe River systems.

This area is on the lands of the Yorta Yorta and Taungurung people, Traditional Owners of the land and waters in the Goulburn and Broken catchments, and the Dja Dja Wurrung, Barapa Barapa and Wamba Wemba people, Traditional Owners of the land and waters in the Campaspe and Loddon catchments.

Flows in the rivers of the GRANT Area are highly regulated by large reservoirs, such as Lake Eildon and Lake Eppalock, and numerous smaller weirs.
Delivery of Commonwealth Environmental Water in the GRANT Area is almost entirely confined to rivers.
However, there are opportunities to deliver small volumes of water to benefit the environment in some wetlands.
The majority of environmental water is delivered to the Goulburn River, with smaller volumes delivered to the Loddon, Campaspe and Broken rivers.

The following code extracts all streams connected to the Goulburn River and its major tributaries from the Murray-connected stream network.

```{r}
#| label: fig-filter_grant_streams
#| echo: true
#| message: false
#| warning: false
#| results: 'hide'
#| fig-cap: "Streams within the Goulburn River and Northern Tributaries (GRANT) area"
#| fig-width: 8
#| fig-height: 6

streams_murray_named <- readRDS("data/murray_networkstream_named.rds")

# Identify Goulburn River segments based on known named segments
goulburn_streams <- streams_murray_named[streams_murray_named$Name %in% c("GOULBURN RIVER", "LODDON RIVER", "BROKEN CREEK", "BROKEN RIVER", "CAMPASPE RIVER"),]

# Get upstream-most segment of each stream in "GOULBURN RIVER", "LODDON RIVER", "BROKEN CREEK", "BROKEN RIVER", "CAMPASPE RIVER":
goulburn_upstream_most <- goulburn_streams %>%
  group_by(Name) %>%
  slice_max(order_by = d2ol, n = 1) %>%
  ungroup()

# find all upstream segments connected to the identified Goulburn River segment
goulburn_ig <- subcomponent(g, as.character(goulburn_upstream_most$From_Node), mode = "in")

goulburn_us <- streams_murray_named %>%
  filter(From_Node %in% names(V(g))[goulburn_ig] |
         To_Node %in% names(V(g))[goulburn_ig])

goulburn_all <- rbind(goulburn_streams, goulburn_us)

# Find unique streams based on hydroID (using base-R to avoid dplyr grouping issues):
goulburn_all <- goulburn_all[!duplicated(goulburn_all$HydroID), ]

# Names of streams in the Goulburn network:
unique(goulburn_all$Name)
GRANT_names <- as.data.frame(table(goulburn_all$Name))

# Extract Murray River segment for context in plots:
muuray_stream <- streams_murray_named[streams_murray_named$Name == "MURRAY RIVER", ]

# Plot streams_murray_connected overlaid on a map of Australia to verify
australia <- rnaturalearth::ne_countries(scale = "medium", returnclass = "sf")
ggplot() +
  geom_sf(data = australia, fill = "grey90") +
  geom_sf(data = muuray_stream, color = "grey40") +
  geom_sf(data = goulburn_all, color = "blue3") +
  # Plot a point for the downstream-most point on the Murray River (murray_ol):
  # geom_sf(data = murray_ol_sf, color = "red", size = 2) +
  coord_sf(xlim = c(142, 149), ylim = c(-39, -34)) +
  theme_minimal() +
  labs(title = "Streams within the Goulburn River and Northern Tributaries (GRANT) area")

saveRDS(goulburn_all, "data/goulburn_networkstream.rds")


```

There are 51 named streams within the Goulburn River and Northern Tributaries (GRANT) area, across 674 stream segments extracted from the Geofabric dataset.
Many of these streams are smaller tributaries surrounding the Lake Eildon reservoir.
The majority of stream segments (n) constitute the Goulburn River (n = 212), Loddon River (n = 107), Broken River (n = 70), Campaspe River (n = 77), and Holland (n = 28) systems.

### 2.3. Extracting catchments within the Murray-Darling Basin

To extract the catchment boundary, we read in the AHGFCatchment layer from the Geofabric dataset and filter for catchments within the Goulburn basin.

The catchment files are very large, and are best filtered at the area-scale using SQL queries when reading in the data.
Attempting to read in the entire AHGFCatchment layer may lead to memory issues if the system is not equipped to handle large datasets (e.g., \<16 GB RAM, such as a laptop or office desktop PC).

```{r}
#| label: extract_grant_catchment_data
#| echo: true
#| message: false
#| warning: false
#| results: 'hide'
#| eval: false

# Explore available layers in the geofabric geodatabase:
st_layers(paste0(geofab_network_path, "SH_Network.gdb/")) # AHGFCatchment

# AHGFCatchment layer contains catchment geometry and attributes
# Inspect field names in AHGFCatchment layer
cat_head <- st_read(
  dsn = paste0(geofab_network_path, "SH_Network.gdb/"),
  layer = "AHGFCatchment",
  query = "SELECT * FROM AHGFCatchment LIMIT 250"
)

# names(cat_head)
# Column headers of interest:
# [1] "StreamName"      "HydroID"    "NextDownID"

# Read in GRANT streams to get list of catchment names:
goulburn_networkstream <- readRDS("data/goulburn_networkstream.rds")
ID_list <- paste0("'", goulburn_networkstream$HydroID[1:3], "'", collapse = ", ")
# Create SQL query to filter catchments by HydroID:
query_string <- paste0("SELECT * FROM AHGFCatchment WHERE HydroID IN (", ID_list, ")")


# read in geofabric catchment data for the goulburn basin:
cat_sf <- sf::st_read(
  dsn = paste0(geofab_network_path, "SH_Network.gdb/"),
  layer = "AHGFCatchment",
  query = "SELECT * FROM AHGFCatchment WHERE StreamName NOT IN ('', ' ')",
  geometry_column = "Shape"
)

grant_cat_sf <- cat_sf[cat_sf$HydroID %in% goulburn_networkstream$DrainID, ]

saveRDS(grant_cat_sf, "data/goulburn_catchments.rds")

# Plot catchments to verify
australia <- rnaturalearth::ne_countries(scale = "medium", returnclass = "sf")
ggplot() +
  geom_sf(data = australia, fill = "grey90") +
  geom_sf(data = grant_cat_sf, fill = "lightblue", color = "blue3", alpha = 0.5) +
  coord_sf(xlim = c(142, 149), ylim = c(-39, -34)) +
  theme_minimal() +
  labs(title = "Catchments within the Goulburn River and Northern Tributaries (GRANT) area")




```

## 3. Creating a naming convention for each stream segment

To facilitate easier identification of stream segments, I have created a naming convention that combines the stream name with a unique segment identifier.
This is based on the d2ol and the name of the stream relating to each segment.
This will help in future analysis and reporting by providing an interpretable identifier for each segment, is robust to automatic additions of new sites, and is also relatable to the Geofabric dataset.

The convention uses the following logic to ensure uniqueness and interpretation:

1 - A three-letter initialism of the stream name (e.g., GOR for Goulburn River), 2 - A single integer corresponding to the Flow-MER river system (e.g., 9 for Goulburn River and Northern Tributaries),\
3 - An integer recording its d2ol rounded to the nearest meter (e.g., 12543 for a segment 12.543 km from the outlet),

The following code implements this naming convention for each stream segment in the Goulburn network and checks for uniqueness across all stream segments.

### 3.1. 3-letter initialism creation

This can be done for every named stream in the Geofabric for consistency.
The following code creates a unique 3-letter initialism for each stream name in the Murray-Darling network (or at least attempts to):

```{r}
#| label: create_stream_naming_convention
#| echo: true
#| message: false
#| warning: false
#| results: 'hide'
#| eval: true

# Read in Goulburn stream network data:
#goulburn_networkstream <- readRDS("data/goulburn_networkstream.rds")

# Create a new column 'sitecode' combining stream name and a unique segment identifier
#goulburn_networkstream$sitecode <- NA

#names <- unique(goulburn_networkstream$Name)
#names_table <- as.data.frame(table(goulburn_networkstream$Name))

murray_networkstream <- readRDS("data/murray_networkstream_named.rds")
names_table <- as.data.frame(table(murray_networkstream$Name))

# Find number of words per stream name:
names_table$word_count <- sapply(as.character(names_table$Var1), function(x) {
  length(unlist(strsplit(x, " ")))
})

# Function to create a three-letter initialism from stream name:
create_initialism <- function(str_name, name_bank) {
  
  # Remove all non-letter and non-space characters from stream name:
  str_name <- gsub("[^A-Za-z ]", "", str_name)

  # Split stream name:  
  str_name_split <- unlist(strsplit(str_name, " "))
  init_repeated = TRUE
  init_attempt = 1
  circuit_breaker = 50000
  exhaustive_sort_start = 26*26
  char1 <- 1
  char2 <- 1
  char3 <- 1
  common_words <- c("THE", "OF", "AND", "CREEK", "BROOK", "STREAM", "CHANNEL", "GULLY", "WATERWAY", "WATERCOURSE", "BRANCH")
  
  while (init_repeated == TRUE){
    
    # Pre-filter names with four or more words:
    if (length(str_name_split) >= 4) {
      # Remove common words like "THE", "OF", "AND":
      str_name_split <- str_name_split[!toupper(str_name_split) %in% common_words]
    }
    
    ### 1. For names with one word:
    if (length(str_name_split) == 1) {
    
      # Take first two letters of the name, then attempt to take a third letter of the name:
      initials <- c(substr(str_name_split[1], 1, 2), substr(str_name_split[1], 2 + init_attempt, 2 + init_attempt))
      initialism <- toupper(paste(initials, collapse = ""))
    }
    
    ### 2. For names with two words:
    if (length(str_name_split) == 2) {
    
      # Name creeks (with more than 3 letters in its name) differently to rivers:
      if (str_name_split[2] == "CREEK" & nchar(str_name_split[1]) > 3){

        # Take first two letters of first word, then attempt to take a third letter of the first word:
        initials <- c(substr(str_name_split[1], 1, 1), substr(str_name_split[1], 1 + init_attempt, 2 + init_attempt))
        initialism <- toupper(paste(initials, collapse = ""))
      } else {
      
        # Take first two letters of first word and first letter of second word:
        initials <- c(substr(str_name_split[1], 1, 1), substr(str_name_split[1], 1 + init_attempt, 1 + init_attempt), substr(str_name_split[2], 1, 1))
        initialism <- toupper(paste(initials, collapse = ""))
      }
    }
  
    ### 3. For names with three or more words:
    if (length(str_name_split) == 3) {
    
      # Take first letter of each word:
      initials <- c(substr(str_name_split[1], 1, 1), substr(str_name_split[2], 1 + init_attempt, 1 + init_attempt), substr(str_name_split[3], 1, 1))
      initialism <- toupper(paste(initials, collapse = ""))
    }  
    
    ### --- Error handling
    
    # If the initialism only has 1 character, add different letters to the end:
    if (nchar(initialism) == 1){
      initials <- c(initialism, LETTERS[char2], LETTERS[char3])
      initialism <- toupper(paste(initials, collapse = ""))
    }
    
    # If the initialism only has 2 characters, add a different letter to the end:
    if (nchar(initialism) == 2){
      initials <- c(initialism, LETTERS[char3])
      initialism <- toupper(paste(initials, collapse = ""))
    }
    
    # If the initialism has more than 3 characters, truncate to 3 characters:
    if (nchar(initialism) > 3){
      initialism <- substr(initialism, 1, 3)
    }
    
    ### If everything has failed so far, force exhaustive sort:
    if (init_attempt > exhaustive_sort_start){
      initials <- LETTERS[c(char1, char2, char3)]
      initialism <- toupper(paste(initials, collapse = ""))
    }
    
    ### 4. Check if initialism already exists in name bank:
    if (initialism %in% name_bank$initialism) {
      
      # Stop if circuit breaker is reached:
      if (init_attempt >= circuit_breaker) {
        stop(paste("Circuit breaker triggered for stream name:", str_name, 
                   "after", circuit_breaker*2, "attempts to create a unique initialism."))
      }
      
      # Add attempt counter:
      init_attempt <- init_attempt + 1
      
      # Reset char if attempt 676 reached (i.e., ZZ):
      if (init_attempt == exhaustive_sort_start){
        # Declare fully randomised attempt:
        print(paste("Exceeded", exhaustive_sort_start, "attempts for stream name:", str_name, 
                      "- forcing exhaustive sort."))
        char2 <- 1
        char3 <- 1
      }
      
      char3 <- char3 + 1
      if (char3 > 26){
        char3 <- 1
        char2 <- char2 + 1
      }
      if (char2 > 26){
        char2 <- 1
        char1 <- char1 + 1
      }
      if (char1 > 26){
        stop(paste("Unable to sort:", str_name))
      }
      # print(paste("Initialism", initialism, "already exists in name bank, trying again..."))
    } else {
      
      # Switch off repeater
      init_repeated = FALSE
      print(paste("Initialism", initialism, "applied to", str_name, "."))
      init_attempt <- 1
    }
  }
  return(initialism)
}

# Initialise a dataframe to store initialisms recursively (i.e., is called within the create_initialism() function):
name_bank <- data.frame(
  stream_name = character(),
  initialism = character(),
  stringsAsFactors = FALSE
)

for (i in 1:nrow(names_table)) {
  stream_name <- as.character(names_table$Var1[i])
  initialism <- create_initialism(as.character(names_table$Var1[i]), name_bank = name_bank)
  
  # Add to name bank:
  name_bank <- rbind(name_bank, data.frame(
    stream_name = stream_name,
    initialism = initialism,
    stringsAsFactors = FALSE
  ))
}

# Check that every initial is unique:
length(unique(name_bank$initialism)) == length(name_bank$initialism) # 0

# Check that all stream names have an initialism with three characters:
length(name_bank$initialism[nchar(name_bank$initialism) != 3]) # 0

# Check that all stream initials are letters:
length(name_bank$stream_name[which(!grepl("^[A-Z]{3}$", name_bank$initialism))]) # 0

# All good!
# Write to file

write.csv(name_bank, "data/murray_stream_name_initialisms.csv", row.names = FALSE)

```

### 3.2. Applying the naming convention to Goulburn stream segments

The final sitecode for each stream segment combines the 3-letter initialism, the Flow-MER river system code (9 for Goulburn River and Northern Tributaries), and the d2ol rounded to the nearest meter, all separated by a hyphen ('-').
This convention requires that each stream segment is at least 1m apart to ensure uniqueness.
The following code applies the naming convention to each stream segment in the Goulburn network:

```{r}
#| label: apply_stream_naming_convention
#| echo: true
#| message: false
#| warning: false
#| results: 'hide'
#| eval: true

# Read in Goulburn stream network data:
goulburn_networkstream <- readRDS("data/goulburn_networkstream.rds")
goulburn_catchments <- readRDS("data/goulburn_catchments.rds")
name_bank <- read.csv("data/murray_stream_name_initialisms.csv")

# Merge name_bank with goulburn_networkstream to get initialisms:
goulburn_networkstream$sitecode <- NA

for (i in 1:nrow(goulburn_networkstream)) {
  stream_name <- goulburn_networkstream$Name[i]
  initialism <- name_bank$initialism[name_bank$stream_name == stream_name]
  
  # Create sitecode:
  d2ol_rounded <- round(goulburn_networkstream$d2ol[i])
  sitecode <- paste0(initialism, "-9-", d2ol_rounded)
  
  goulburn_networkstream$sitecode[i] <- sitecode
}

# Check for uniqueness of sitecodes:
length(unique(goulburn_networkstream$sitecode)) == nrow(goulburn_networkstream) # TRUE

# Merge geometries for sites less than 1m apart (in base R):
goulburn_networkstream <- goulburn_networkstream[order(goulburn_networkstream$d2ol, decreasing = TRUE),]
goulburn_networkstream_agg <- aggregate(goulburn_networkstream["sitecode"], by = list(goulburn_networkstream$sitecode), FUN = function(x) x[1], do_union = TRUE)

# drop sf from goulburn_networkstream
goulburn_networkstream <- st_set_geometry(goulburn_networkstream, NULL)
class(goulburn_networkstream) <- "data.frame"

# Merge goulburn_networkstream onto goulburn_networkstream_agg to get other attributes back:
goulburn_networkstream_merge <- merge(goulburn_networkstream_agg, goulburn_networkstream, by = "sitecode", all.x = TRUE)

# Merge catchment geometries based on sitecode:


```

## 4. References
